### 2024-02-29 Appendix.Rmd
Edits minor revision BRM

```{r echo = FALSE, message = FALSE, fig.width=11.3, fig.height=10.4, fig.align = 'center',fig.cap = "Estimated statistical power for recovering the data-generating model (y-axis) as a function of sample size (x-axis) depending on (A) loading strength for lower versus higher inter-factor correlation and (B) inter-factor correlation for lower versus higher factor loadings. * denotes results from the original analyses in the main manuscript"}

>>>>

```{r echo = FALSE, message = FALSE, fig.width=11.3, fig.height=10.4, fig.align = 'center',fig.cap = "Estimated statistical power for correctly selecting the data-generating model (y-axis) as a function of sample size (x-axis) depending on (A) loading strength for lower versus higher inter-factor correlation and (B) inter-factor correlation for lower versus higher factor loadings. * denotes results from the original analyses in the main manuscript"}

### 2024-02-29 manuscript.Rmd
Edits minor revision BRM

abstract: |
  Determining the compositional structure and dimensionality of psychological constructs lies at the heart of many research questions in developmental science. Structural equation modelling (SEM) provides a versatile framework for formalizing and estimating the relationships among multiple latent constructs. While the flexibility of SEM can accommodate many complex assumptions on the underlying structure of psychological constructs, it makes a-priori estimation of statistical power and required sample size challenging. This difficulty is magnified when comparing non-nested SEMs, which prevents the use of the Likelihood Ratio test. Sample size estimates for SEM model fit comparisons typically rely on generic rules of thumb. Such heuristics can be misleading because statistical power in SEM depends on a variety of model properties. Here, we demonstrate a Monte Carlo simulation approach for estimating a-priori statistical power for non-nested model comparisons in an SEM framework. We provide a step-by-step guide to this approach based on an example from our memory development research in children.

@@

In an SEM framework, theoretical constructs of interest are operationalized as latent variables that capture the underlying commonalities across a set of (measured) manifest variables. Manifest variables can directly be observed (such as items in a questionnaire or indices derived from a behavioral task) and are commonly illustrated as squares in graphical depictions of SEM. Latent variables represent hypothetical constructs that are not directly observable (such as a latent factor of intelligence or memory), but are inferred from a selection of manifest variables. In graphical depictions of SEM, latent variables are commonly shown as circles. The strength of a loading from a manifest onto a latent variable represents the extent to which the observed variance is accounted for by the latent factor. In other words, the loading indicates how well a given manifest variable captures the latent construct.

The relationship among manifest and latent variables can be described in a mathematical model, which specifies all assumed parameters. Two models M1 and M2 are referred to as nested, if the parameter space in a more restrictive model M2 represents a subspace of the parameter space of the more general model M1  [@bentler_1980]. This means that the two models only differ with regards to the specification of one or multiple parameters. Nesting of models is usually achieved via constraining free parameters from M1 to equality or to known constants. The concept of nesting represents a convenient characteristic when comparing competing models, as it allows the use of a Likelihood Ratio test to evaluate their relative model fits [@satorra_1985]. When comparing non-nested models, researchers usually have to revert to model fit indices that indicate how well a given model describes the data, such as the Bayesian Information Criterion [BIC, @raftery_1986; @schwarz_1978], the Akaike Information Criterion [AIC, @akaike_1974], the Root Mean Square Error of Approximation [RMSEA, @steiger_2016], and the Comparative Fit Index [CFI, @bentler_1990]. Such indices can guide heuristic decisions that do not hold any statistical guarantees (e.g., a pre-specified Type-I error level).

@@

Various rules of thumb on required sample size in SEM have been recommended, including setting an absolute minimum of observations across the board  [@boomsma_1985] and adjusting to the model complexity [i.e., setting a number of observations per estimated parameter, @bentler_1987]. Unfortunately, such heuristics can be misleading, as statistical power in SEM is heavily influenced by parameters beyond the number of indicators per latent construct, e.g. the magnitude of factor loadings [@wolf_2013]. Further, determining the required sample size for a given study depends on the research question. One common utility of SEM is to test whether a given effect in a model exceeds a specific threshold, such as testing whether the correlation of latent factors exceeds zero, or whether a specific parameter estimate differs between groups (i.e., target effect). Another common utility is to determine how well a given model describes the data, and/or whether one model describes the data better than a competing model (i.e., model comparison). These different research questions require different types of power: the power to detect a target effect vs. the power to detect model misspecification [@wang_2021]. Further, theoretical assumptions on competing models can lead to the necessity to compare non-nested models, which prevents the use of regular $\chi^2$-based power estimates and therefore requires alternative approaches to determining statistical power for SEM.

@@

### 1.2.2	Statistical Power for Model Comparison 
A second question that is of interest for many researchers – especially in developmental cognitive neuroscience – pertains to identifying the one theoretical model (from a set of competing models) that best explains the given data [@henson_2016; @miller_2012; @nyberg_1994]. In the case of competing models that are nested, the analytical approach to this question (and therefore also considerations on statistical power) can be addressed analogously to the procedure outlined above [see @jobst_2021 for a step-by step tutorial]. However, comparing non-nested models prohibits the use of regular $\chi^2$-based statistics and therefore poses additional methodological challenges in determining the ‘best’ model. Nevertheless, researchers should strive for a methodological approach that matches the theoretical assumptions, rather than vice versa, that is, moving away from well-grounded theoretical considerations to meet methodological constraints. In cases where analytical solutions to questions of statistical properties like statistical power are not available, randomization-based techniques offer an excellent alternative [@efron_1994]. In particular, Monte Carlo simulations have proven useful to bridging this gap [@muthen_2002, see Section 2.1 for more details on Monte Carlo simulations].

@@

Which of these views best explain memory abilities across early development? This question requires adequate statistical methods that can adjudicate among multiple competing ideas on the underlying structure of memory. The simulation in the following section of this paper will address methodological challenges in study planning when aiming to compare three hypothetical models: a unitary model (Model 1), a bi-partite model (Model 2) and a tri-partite model (Model 3) on the compositional structure of memory (Figure 1). Specifically, we will focus on the issue of estimating statistical power for this model comparison. Here, we address the impact of two main aspects in particular: (i) the separability of the theoretical constructs, i.e. inter-relatedness of the latent factors in the competing models and (ii) the reliability of the measures, i.e. the loadings of the manifest variables. Details on the theoretical basis and the selection of methodological indicators for each model can be found in @buchberger_2022. Important for this context is the fact that our specification of Models 2 and 3 results in a non-nested model comparison (due to the cross-loadings for the latent factors pattern separation and pattern completion and different allocation of the indicator Task 10). Therefore a-priori estimations of statistical power call for a simulation-based approach.

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in Section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model among the three competing models, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (Section 3.1) and the reliability of the manifest variables (Section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different conditions of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (Section 3.3).

@@

### 2.1.3	Step 3: Generate 
Next, we define the mechanism to generate the synthetic data. An easy way to generating synthetic data within an SEM framework in R is the simulateData() function available in the lavaan package [@rosseel_2012]. In the example on memory composition we simulate multivariate normal data, as the indicators represent continuous values of task performance from the memory tasks. However, the simulation approach presented here can of course be extended to dichotomous or polytomous data for other use cases. Within the generate function, all parameters that are predetermined across iterations can be directly entered into the generating model syntax. However, all parameters that are supposed to be varied between different iterations of the simulation need to be accessed through the condition argument that is returned through the createDesign() function in step 2. Given our interests in the effect of loading strength and strength of covariance between the latent factors, we define these parameters in step 2 (Design) such that they can now be called via the condition argument in the generate function: 
```{r generate, echo = TRUE}
```

@@

# 3. Results: Power Estimates for Investigating the Compositional Structure of Memory

@@

In the example of comparing competing models of memory structure, we first examined the separability of the theoretical models by investigating the influence of the latent factors’ correlations on statistical power estimates. In our example simulation, we investigated the statistical power to retrieve Model 3 as the best fitting model, when it indeed generated the data, while varying sample sizes from 50 to 200 and factor inter-correlations from 0.1 to 0.9. In this case, we restricted the simulated correlations between the latent factors to positive values, as there is no indication in the literature to suspect a negative relation among these memory capacities [e.g., @tucker-drob_2019]. Figure 3A displays the obtained power curves illustrating the estimated statistical power dependent on the total sample size and the factor inter-correlation. For this estimate, the loading of the manifest variables was fixed at 0.7 and the inter-correlation between all latent factors were varied simultaneously.  
As expected, statistical power increased with increasing sample size. However, the separability of the latent factors mattered: The higher the correlations between factors, the less able we were to distinguish multi-factor models from a one-factor model. That is, the smaller the difference between competing models, the greater the sample size that would be required to dissociate between them with adequate power. For very high levels of inter-correlation (0.8 and higher) the increase of statistical power with increasing sample size was remarkably lower than for lower levels of inter-correlation (0.7 and below). When the inter-factor correlation was 0.9, not even a larger sample size (*N* =200) achieved an adequate level of power of 80%. These results are not surprising given that increasing the inter-correlations between the latent factors renders the models more similar to each other, and thus makes their distinction more challenging. For moderate levels of inter-correlation (<= 0.7) a statistical power of 95% was reached for *N* = 125.

@@

```{r, echo = FALSE, fig.width=11.3, fig.height=4.3, fig.align = 'center', fig.cap = "Simulation results: Estimated statistical power for recovering the data-generating model, i.e. Model 3 (y-axis) as a function of sample size (x-axis) with varying the factor inter-correlations when the loading strength was fixed to 0.7 (A) $;$ and with varying the loading strength when the factor inter-correlation was fixed to 0.3 (B). For each combination of parameters, 1,000 datasets were simulated. The red dashed horizontal line positioned at 95\\% discovery rate reflects the cut-off for identifying the appropriate sample size. The syntax to generate these plots with ggplot() can be found in the supplementary R Code on GitHub."}

@@

## 3.3	How can Simulations Inform us About Type-I Error Rates?
The main part of this tutorial has focused on estimating statistical power that is required to reject a null hypothesis when it is indeed false (Type-II error). However, it is necessary to also consider the reverse — rejecting a null hypothesis that is in fact true (Type-I error). In our example, Type-I error would be committed by misidentifying Model 3 as the best fitting model when in reality, the data were generated from either Models 1 or 2. We can leverage the simulation approach to quantify Type-I error rates dependent on the different simulation conditions (e.g., inter-factor correlations, factor loadings). We conducted an analogous analysis to the one outlined above (Sections 2.1.1. – 2.1.6), with the difference that Models 1 and 2 are defined as the generating population models (step 3, Section 2.1.3). The outcomes of this analysis show the probability of erroneously identifying an overly complex model, when the more simplistic models generated the data (see Figure 4).

@@

SEM offers a powerful tool for modeling multivariate relationships between psychological constructs. Maximizing its utility in psychological sciences hinges on the integration of its complex characteristics with power estimation – another key tenet of scientific rigor. Our work exemplifies a roadmap for navigating some of the challenges in harmonizing these two efforts by implementing a-priori power estimation within an SEM framework. In the case of information-theoretic model comparisons for non-nested models, statistical power cannot be determined analytically, i.e., model comparisons cannot be performed by using $\chi^2$-based statistics. Monte Carlo simulations provide a useful tool to circumvent this methodological challenge [@muthen_2002], allowing researchers to derive informed decisions on sample size planning to ensure sufficient statistical power and prevent uninformative or misleading research [@button_2013]. In this study, we illustrate a step-by-step approach to implementing a Monte Carlo simulation in the statistical programming language R to estimate the required sample size for model comparisons of non-nested models. The complete code is available on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM) [@zenodo-release] to yield a starting point for researchers who want to implement their own Monte Carlo power simulations.

@@

Further, we have presented an approach that leverages a combination of absolute and relative model fit parameters to adjudicate between competing models. While such parameter-based decisions are well established in the field, they may result in choosing one model over the other, even in cases when differences in model fit parameters are very small, as one can never conclude that two models fit equally well or that there is too little data to make a decision with enough confidence. An alternative approach to this criterion for model selection originates from a theory by @vuong_1989, who suggested likelihood ratio tests for non-nested models. Recent advances that build on this idea have postulated a framework that allows to test hypotheses of model distinguishability and difference in fit [ @levy_2007; @levy_2011], which can be applied to non-nested SEMs and easily implemented via new software packages in R [ nonnest2, @merkle_2016]. The resulting recommendations for the comparison of non-nested SEMs posit a stepwise procedure to test whether competing non-nested models are distinguishable in a given population and, if yes, whether one fits the data significantly better than the other. Importantly, this method derives interval estimates for differences in non-nested information criteria [ @merkle_2016 ]. Thus, this approach does not only allow to conclude that one or the other model fits the data better, but also allows for the possibility to conclude that there is insufficient evidence to determine which of two models fits a given dataset better. This stepwise procedure can be implemented as a model selection criterion in a power simulation like the one presented above. To integrate model selection via a likelihood ratio test for non-nested models into the current simulation, one would need to adapt the summarize function accordingly, such that it identifies model 3 as the best fitting model only if a) the competing models are distinguishable on the simulated data set of a given iteration, b) the Likelihood ratio test for non-nested models indicates in a significantly better fit for model 3 compared to models 1 and 2 and c) model 3 reaches the cutoffs for absolute model fit, as defined in the previous simulation. The vuongtest() function from the nonnest2 package [@merkle_2014] can be a helpful tool for integrating these tests into the simulation. While a systematic investigation of the interplay of indicator reliability, separability of latent factors and statistical power for model selection via model fit parameters versus the LRT for non-nested models is beyond the scope of this tutorial, it is worth noting that such an alternative approach may yield better model recovery in certain cases. The conditions under which model selection via the LRT for non-nested models provides better estimates of statistical power are thus an interesting avenue for future research.

@@

In sum, we argue that sample size planning in the case of non-nested SEM comparisons requires a simulation-based approach. In particular, we highlighted that the separability of theoretical constructs, as well as the reliability of the measures, have a major impact on estimates of statistical power. To this end we hope that this tutorial advances the use of simulation-based approaches to estimating statistical power in cases where analytical approaches do not suffice.

>>>>

abstract: |
  Determining the compositional structure and dimensionality of psychological constructs lies at the heart of many research questions in developmental science. Structural equation modelling (SEM) provides a versatile framework for formalizing and estimating the relationships among multiple latent constructs. While the flexibility of SEM can accommodate many complex assumptions on the underlying structure of psychological constructs, it makes a-priori estimation of statistical power and required sample size challenging. This difficulty is magnified when comparing non-nested SEMs, which prevents the use of traditional Likelihood Ratio tests. Sample size estimates for SEM model fit comparisons typically rely on generic rules of thumb. Such heuristics can be misleading because statistical power in SEM depends on a variety of model properties. Here, we demonstrate a Monte Carlo simulation approach for estimating a-priori statistical power for model selection when comparing non-nested models in an SEM framework. We provide a step-by-step guide to this approach based on an example from our memory development research in children.

@@

In an SEM framework, theoretical constructs of interest are operationalized as latent variables that capture the underlying commonalities across a set of (measured) manifest variables. Manifest variables can directly be observed (such as items in a questionnaire or indices derived from a behavioral task) and are commonly illustrated as squares in graphical depictions of SEM. Latent variables represent hypothetical constructs that are not directly observable (such as a latent factor of intelligence or memory), but are inferred from a selection of manifest variables. In graphical depictions of SEM, latent variables are commonly shown as circles. The strength of a loading, that is, a path from a latent variable to an observed variable, represents the extent to which the observed variance is accounted for by the latent factor. In other words, the loading indicates how well a given manifest variable captures the latent construct.

The relationship among manifest and latent variables can be described in a mathematical model, which specifies all assumed parameters. Two models M1 and M2 are referred to as nested, if the parameter space in a more restrictive model M2 represents a subspace of the parameter space of the more general model M1  [@bentler_1980]. This means that the two models only differ with regards to the specification of one or multiple parameters. Nesting of models is usually achieved via constraining free parameters from M1 to equality or to known constants. The concept of nesting represents a convenient characteristic when comparing competing models, as it allows the use of a Likelihood Ratio test to evaluate their relative model fits [@satorra_1985]. When comparing non-nested models, researchers usually revert to the comparison of model fit indices that indicate how well a given model describes the data, such as the Bayesian Information Criterion [BIC, @raftery_1986; @schwarz_1978], the Akaike Information Criterion [AIC, @akaike_1974], the Root Mean Square Error of Approximation [RMSEA, @steiger_2016], and the Comparative Fit Index [CFI, @bentler_1990, but see section 4 for a discussion of recently proposed non-nested Likelihood Ratio tests ]. Such indices can guide heuristic model selection, but do not allow for any statistical guarantees (e.g., a pre-specified Type-I error level).

@@

Various rules of thumb on required sample size in SEM have been recommended, including setting an absolute minimum of observations across the board  [@boomsma_1985] and adjusting to the model complexity [i.e., setting a number of observations per estimated parameter, @bentler_1987]. Unfortunately, such heuristics can be misleading, as statistical power in SEM is heavily influenced by parameters beyond the number of indicators per latent construct, e.g. the magnitude of factor loadings [@wolf_2013]. Further, determining the required sample size for a given study depends on the research question. One common utility of SEM is to test whether a given effect in a model exceeds a specific threshold, such as testing whether the correlation of latent factors exceeds zero, or whether a specific parameter estimate differs between groups (i.e., target effect). Another common utility is to determine how well a given model describes the data, and/or whether one model describes the data better than a competing model (i.e., model comparison). These different research questions require different types of power: the power to detect a target effect vs. the power to detect model misspecification [@wang_2021]. Further, theoretical assumptions on competing models can lead to the necessity to compare non-nested models, which prevents the use of traditional $\chi^2$-based power estimates and therefore requires alternative approaches to determining statistical power for SEM (see section 4 for a more detailed discussion of recently proposed non-nested Likelihood Ratio tests).

@@

### 1.2.2	Statistical Power for Model Comparison 
A second question that is of interest for many researchers – especially in developmental cognitive neuroscience – pertains to identifying the one theoretical model (from a set of competing models) that best explains the given data [@henson_2016; @miller_2012; @nyberg_1994]. In the case of competing models that are nested, the analytical approach to this question (and therefore also considerations on statistical power) can be addressed analogously to the procedure outlined above [see @jobst_2021 for a step-by step tutorial]. However, comparing non-nested models prohibits the use of traditional $\chi^2$-based statistics and therefore poses additional methodological challenges in determining the ‘best’ model. Nevertheless, researchers should strive for a methodological approach that matches the theoretical assumptions, rather than vice versa, that is, moving away from well-grounded theoretical considerations to meet methodological constraints. In cases where analytical solutions to questions of statistical properties like statistical power are not available, randomization-based techniques offer an excellent alternative [@efron_1994]. In particular, Monte Carlo simulations have proven useful to bridging this gap [@muthen_2002, see Section 2.1 for more details on Monte Carlo simulations].

@@

Which of these views best explain memory abilities across early development? This question requires adequate statistical methods that can adjudicate among multiple competing ideas on the underlying structure of memory. The simulation in the following section of this paper will address methodological challenges in study planning when aiming to compare three hypothetical models: a unitary model (Model 1), a bi-partite model (Model 2) and a tri-partite model (Model 3) on the compositional structure of memory (Figure 1). Specifically, we will focus on the issue of estimating statistical power for this model comparison. Here, we address the impact of two main aspects in particular: (i) the separability of the theoretical constructs, i.e. inter-relatedness of the latent factors in the competing models and (ii) the reliability of the measures, i.e. the loadings of the manifest variables. Details on the theoretical basis and the selection of methodological indicators for each model can be found in @buchberger_2022. Important for this context is the fact that our specification of Models 2 and 3 results in a non-nested model comparison (due to the cross-loadings for the latent factors pattern separation and pattern completion and different allocation of the indicator Task 10). Therefore a-priori estimations of statistical power for selecting the correct model call for a simulation-based approach.

# 2. Methods
In the following, we demonstrate how to conduct a simulation-based power analysis for non-nested model comparisons. Here, we refer to statistical power as the probability of correctly selecting a model given a selection of candidate models. In a hypothesis testing framework, the alternative hypothesis corresponds to a given model being the true model. Statistical power then reflects the correct selection rate of a model and the type-I error corresponds to the probability of incorrectly selecting that model when one of the other candidates is the true model. For the sake of illustration, we will refer to the example on the compositional structure of memory outlined in Section 1.3 throughout. For this example, we aim to derive estimates of required sample size to identify the tri-partite model as the best fitting model among the three competing models, if it indeed generated the data. In a step-by-step guide, we therefore show how to specify the competing models in R, how to set up different design conditions for the simulation, how to define the functions to generate, analyze and summarize synthetic data and how to execute the simulation. Based on two separate simulations, we show how the separability of theoretical constructs (Section 3.1) and the reliability of the manifest variables (Section 3.2) impacts estimates of statistical power for model comparisons by varying inter-factor correlations and factor loadings across different conditions of the simulation. Finally, we employ the same simulation-based approach to investigate Type-I error rates, that is erroneously identifying the tri-partite model as the best fitting model if the bi-partite or unitary model generated the data (Section 3.3).

@@

### 2.1.3	Step 3: Generate 
Next, we define the mechanism to generate the synthetic data. An easy way to generating synthetic data within an SEM framework in R is the simulateData() function available in the lavaan package [@rosseel_2012]. In the example on memory composition we simulate multivariate normal data, as the indicators represent continuous values of task performance from the memory tasks. However, the simulation approach presented here can of course be extended to dichotomous or polytomous data for other use cases. Within the generate function, all parameters that are predetermined across iterations can be directly entered into the generating model syntax. However, all parameters that are supposed to be varied between different iterations of the simulation need to be accessed through the condition argument that is returned through the createDesign() function in step 2. It can be useful to examine the resulting datasets prior to running the simulation to assess whether the data generating mechanism returns realistic datasets (e.g. regarding the variability of the data, extreme values), to allow for refining the generate function accordingly [similar to prior predictive checks as often suggested in the context of Bayesian approaches, @gelman_2020]. Given our interests in the effect of loading strength and strength of covariance between the latent factors, we define these parameters in step 2 (Design) such that they can now be called via the condition argument in the generate function: 
```{r generate, echo = TRUE}
```

@@

# 3. Results: Statistical Power Estimates for Investigating the Compositional Structure of Memory

@@

In the example of comparing competing models of memory structure, we first examined the separability of the theoretical models by investigating the influence of the latent factors’ correlations on statistical power estimates. In our example simulation, we investigated the statistical power to retrieve Model 3 as the best fitting model, when it indeed generated the data, while varying sample sizes from 50 to 200 and factor inter-correlations from 0.1 to 0.9. In this case, we restricted the simulated correlations between the latent factors to positive values, as there is no indication in the literature to suspect a negative relation among these memory capacities [e.g., @tucker-drob_2019]. Figure 3A displays the obtained power curves illustrating the estimated statistical power dependent on the total sample size and the factor inter-correlation. For this estimate, the loading of the manifest variables was fixed at 0.7 and the inter-correlation between all latent factors were varied simultaneously.  
As expected, statistical power increased with increasing sample size. However, the separability of the latent factors mattered: The higher the correlations between factors, the less able we were to distinguish multi-factor models from a one-factor model. That is, the smaller the difference between competing models, the greater the sample size that would be required to dissociate between them with adequate power. For very high levels of inter-correlation (0.8 and higher) the increase of statistical power with increasing sample size was remarkably lower than for lower levels of inter-correlation (0.7 and below). When the inter-factor correlation was 0.9, not even a larger sample size (*N* =200) achieved an adequate level of power of 80%. These results are not surprising given that increasing the inter-correlations between the latent factors renders the models more similar to each other or - in the case of correlations approaching 1 - identical, making their distinction impossible. For moderate levels of inter-correlation (<= 0.7) a statistical power of 95% was reached for *N* = 125.

@@

```{r, echo = FALSE, fig.width=11.3, fig.height=4.3, fig.align = 'center', fig.cap = "Simulation results: Estimated statistical power for correctly selecting the data-generating model, i.e. Model 3 (y-axis) as a function of sample size (x-axis) with varying the factor inter-correlations when the loading strength was fixed to 0.7 (A) $;$ and with varying the loading strength when the factor inter-correlation was fixed to 0.3 (B). For each combination of parameters, 1,000 datasets were simulated. The red dashed horizontal line positioned at 95\\% discovery rate reflects the cut-off for identifying the appropriate sample size. The syntax to generate these plots with ggplot() can be found in the supplementary R Code on GitHub."}

@@

## 3.3	How can Simulations Inform us About Type-I Error Rates?
The main part of this tutorial has focused on estimating statistical power that is required to reject a null hypothesis when it is indeed false. However, it is necessary to also consider the reverse — rejecting a null hypothesis that is in fact true (Type-I error). In our example, Type-I error would be committed by misidentifying Model 3 as the best fitting model when in reality, the data were generated from either Models 1 or 2. We can leverage the simulation approach to quantify Type-I error rates dependent on the different simulation conditions (e.g., inter-factor correlations, factor loadings). We conducted an analogous analysis to the one outlined above (Sections 2.1.1. – 2.1.6), with the difference that Models 1 and 2 are defined as the generating population models (step 3, Section 2.1.3). The outcomes of this analysis show the probability of erroneously identifying an overly complex model, when the more simplistic models generated the data (see Figure 4).

@@

SEM offers a powerful tool for modeling multivariate relationships between psychological constructs. Maximizing its utility in psychological sciences hinges on the integration of its complex characteristics with power estimation – another key tenet of scientific rigor. Our work exemplifies a roadmap for navigating some of the challenges in harmonizing these two efforts by implementing a-priori power estimation within an SEM framework. In the case of information-theoretic model comparisons for non-nested models, statistical power cannot be determined based on traditional $\chi^2$-based statistics. Monte Carlo simulations provide a useful tool to circumvent this methodological challenge [@muthen_2002], allowing researchers to derive informed decisions on sample size planning to ensure sufficient statistical power and prevent uninformative or misleading research [@button_2013]. In this study, we illustrate a step-by-step approach to implementing a Monte Carlo simulation in the statistical programming language R to estimate the required sample size for model comparisons of non-nested models. The complete code is available on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM) [@zenodo-release] to yield a starting point for researchers who want to implement their own Monte Carlo power simulations.

@@

Further, we have presented an approach that leverages a combination of absolute and relative model fit parameters to adjudicate between competing models. While such parameter-based decisions are well established in the field, they may result in choosing one model over the other, even in cases when differences in model fit parameters are very small, as one can never conclude that two models fit equally well or that there is too little data to make a decision with enough confidence. An alternative approach to this criterion for model selection originates from a theory by @vuong_1989, who suggested likelihood ratio tests for non-nested models. Recent advances that build on this idea have postulated a framework that allows to test hypotheses of model distinguishability and difference in fit [ @levy_2007; @levy_2011], which can be applied to non-nested SEMs and easily implemented via new software packages in R [ nonnest2, @merkle_2016]. The resulting recommendations for the comparison of non-nested SEMs posit a stepwise procedure to test whether competing non-nested models are distinguishable in a given population and, if yes, whether one fits the data significantly better than the other. Importantly, this method derives interval estimates for differences in non-nested information criteria [ @merkle_2016 ]. Thus, this approach does not only allow to conclude that one or the other model fits the data better, but also allows for the possibility to conclude that there is insufficient evidence to determine which of two models fits a given dataset better. This stepwise procedure can be implemented as a model selection criterion in a power simulation like the one presented above. To integrate model selection via a likelihood ratio test for non-nested models into the current simulation, one would need to adapt the summarize function accordingly, such that it identifies model 3 as the best fitting model only if a) the competing models are distinguishable on the simulated data set of a given iteration, b) the Likelihood Ratio test for non-nested models indicates a significantly better fit for model 3 compared to models 1 and 2 and c) model 3 reaches the cutoffs for absolute model fit, as defined in the previous simulation. The vuongtest() function from the nonnest2 package [@merkle_2014] can be a helpful tool for integrating these tests into the simulation. While a systematic investigation of the interplay of indicator reliability, separability of latent factors and statistical power for model selection via model fit parameters versus the Likelihood Ratio test for non-nested models is beyond the scope of this tutorial, it is worth noting that such an alternative approach may yield better model recovery in certain cases. The conditions under which model selection via the Likelihood Ratio test for non-nested models provides better estimates of statistical power are thus an interesting avenue for future research.

@@

In sum, we argue that sample size planning in the case of non-nested SEM comparisons can be achieved via a simulation-based approach. In particular, we highlighted that the separability of theoretical constructs, as well as the reliability of the measures, have a major impact on estimates of statistical power. To this end we hope that this tutorial advances the use of simulation-based approaches to estimating statistical power for model selection when comparing non-nested models in an SEM framework.
